The abundance of Bullet-Groups in $\Lambda$CDM

### 2014-03-20 paper/ms.tex
new version

We estimate the expected distribution of displacements between the two
dominant dark matter density peaks and between baryons and dark matter
in halos simulated within a full cosmological context in $\Lambda$CDM.
We use as a benchmark the observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated
baryonic component. We extend to lower masses previous computational results by
studying two halo samples: groups and clusters. We find that $(50 \pm
10)\%$ of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$186 \pm 30$\hkpc as observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $(70 \pm 10)\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1$ to
$1.0\%$ of the groups should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for clusters this fraction rises to $(7\pm 3)\%$, consistent
with previous studies of dark matter to baryon separations. In terms
of absolute number densities we find that the expected value for
groups similar to \bullg\ is tree times larger than the expected value
for clusters. This extension of the theoretical predictions and
observational results towards low mass bi-modal dark matter
configurations opens up the possibility for a new statistical test of
$\Lambda$CDM.   
\end{abstract}

@@

In this letter we work with three quantities that could be inferred
from observations of Bullet-like systems. The projected distance
between two dominant DM clumps, $d_{\rm 2D}$; the projected distance
between the DM-baryonic clumps, $d_{\rm 2D}^{\rm bar}$; and the mass
ration associated to the two dominant DM clumps.

From the simulation point of view, the first quantity can be
translated into the 2D projected values of $||\vec{r}||$ and its value
relative to the virial radius $D_{\rm   off}= ||\vec{r}||_{\rm
  2D}/R_{\rm vir}$. The second quantity, the projected DM-baryon
distance, is not directly available from a DM-only simulation but can
be estimated from the data.  The last quantity, the mass ratio, can be
approximated by ratio of the circular velocities of the two clumps
$V_{\rm c,sub}/V_{\rm  c,host}$ which should be equal to the square
root of the mass ratio.

@@

\subsection{Displacements and Relative Circular Velocities}
\label{fig:displacement}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_2.eps} 
\end{center}
\caption{2D histogram in the plane $V_{\rm c,sub}/V_{\rm
    c,host}$-$d_{2D}$. The left panel corresponds to clusters and the
  right panel to groups. The star with error bars corresponds to
  \bullg\ data reported from \citet{2013A&A...552A..80M} and
  \citet{Gastaldello}. The data used to construct the histograms
  integrates the objects at all redshifts. }
\label{fig:mass_displacement}
\end{figure*}

The main result of this paper is summarized in Figure
\ref{fig:displacement}, it presents the integrated
probability distribution for the displacement between the center of
the host halo and its dominant sub-halo. The left panel shows
the displacement in physical units and the right panel as a fraction
of the virial radius of the host halo.

Figure \ref{fig:displacement} shows the results for two different
populations at all redshifts. The panel with the projected 2D physical displacements also shows a
vertical stripe with the estimated displacement for the Bullet-group reported by
\cite{Gastaldello}. In the group sample we see that a fraction of
$40\%$ to $60\%$  should present a displacement equal than the
estimate for \bullg; in the cluster sample this fraction increases to
$70\%$-$80\%$. The uncertainties in this estimates are derived from
the uncertainties in the displacement measurement for \bullg.This
fraction is naturally higher in more massive systems because they are
larger in size. Normalizing the displacements by the virial radius,
right panel Figure \ref{fig:displacement}, we see that the
distribution is the same regardless of the sample and the redshift.

Figure \ref{fig:mass_displacement} shows 2D histograms in a plane
defined by the ratio of the two circular velocities $V_{\rm c,
  sub}/V_{\rm c, host}$ and the projected 2D physical displacements;
quantities that can be constrained by observations.

To construct this figure we co-add all the halos in the sample (left,
clusters; right, groups) at all redshifts. We stack the data because
we do not observe any strong redshift dependence. This allows
us to increase the signal in each bin. We overplot a star with error
bars that represents the observational constraints for the system
\bullg\ using the fraction in velocity dispersion in the line-of-sigth
of the group SL2S SJ08544-0121 ($\sigma_{host}=341^{+43}_{-109}$\kms and
$\sigma_{sub}=185^{+30}_{-62}$\kms) reported by
\citet{2013A&A...552A..80M} and the DM displacement inferred from the
data presented by \cite{Gastaldello}.

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_3.eps}
\end{center}
\caption{2D histograms in the plane $\mu$-$D_{\rm off}$. The left
  panel corresponds to clusters and the right panel to groups. The data
  used to construct this histogram includes the halos at all redshifts.}
\label{fig:geometry}
\end{figure*}

@@

Figure \ref{fig:geometry} presents the geometry of the bullet groups using the 
variables $\mu$ and $D_{\rm off}$. The first evident feature is that
most of the configurations have $|\mu|>0.9$ ($\theta\leq 30^{\circ}$),
meaning that most of the collisions can be described as a head-on
encounter while only a minority with $|\mu|<0.9$ have grazing
trajectories. For the pairs on radial trajectories there are three
regions of interest in this plane that describe different stages in
the collision, assuming that the sub-halo merges (or falls below the
BDM detection threshold) right at its second pass through the center
of the host halo \citep{Poole2006}.

The first region has $\mu\approx-1$ and $D_{\rm off}>0.6$, which
locates the systems where a head-on collision has just started. The
sub-halo is close to the boundary of the host halo and is
infalling. The second region has $\mu\approx 1$ and $D_{\rm off}<0.6$;
at this stage the collision continues after the first crossing of the
host's center, the low number of halos with radial infalling velocities
and displacements $D_{\rm off}>0.6$ suggest that this the maximum
range of radii for the apogee.  The third region corresponds to
$\mu\approx-1$ and $D_{\rm off}<0.6$ that is the the
secondary infall after apogee. We use this sequence in the next
subsection to estimate the expected displacement between baryons and
the dominant DM clump.

\subsection{Displacement between Dark Matter and Baryons}
\label{sec:baryonic_displacements}

Strictly speaking, the results we have derived so far apply to multi-modal
groups and their expected separation between the two dominant dark
matter clumps. However, a non-zero DM-DM displacement cannot be
interpreted as having a corresponding non-zero DM-baryon
displacement. Given that we use a DM only simulation, we have to find
a way to estimate DM-baryon displacements from the available information.

This estimate is based on the different collision stages described at
the end of the previous sub-section.  For instance, in systems where
the halo is starting to fall into the host ($\mu\approx-1$, $D_{\rm
  off}>0.6$) the DM-baryon displacement should be equal to
zero. However, when the  sub-structure has already passed through the
center of the host halo, i.e. cases where $\mu\approx 1$, the
DM-baryon displacement should be different from zero.

To estimate the DM-baryon displacement we  work under the following
hypothesis. First, we consider that systems with $|\mu|<0.9$ have a
baryonic displacement, $d_{2D}^{\rm bar}$, equal to zero. This means
that only head-on encounters produce a displacement. Second, we
consider that all systems with infalling velocities $\mu<-0.9$ and
large displacements $D_{\rm off}>0.6$ also have baryonic displacements
equal to zero. Third, for all the other cases we estimate the
displacement between the baryons and the dominant DM peak by
$d_{2D}^{\rm bar} = X_{\rm off} R_{\rm   vir}$, where $X_{\rm   off}$
is the offset computed between the minium of potential and the center
of mass for each host halo computed from all the matter inside the
virial radius.

This simplified model does not take into account that there is a
fraction of halos with $\mu<-0.9$ and $D_{\rm off}<0.6$ for which the
collision has not started and should have $d_{2D}^{\rm bar}=0$. A detailed
modeling of this fraction requires the study of the complete merger
tree of the halo and sub-halo, a study beyond the scope of this
Letter. Instead we caution the reader that the derived fraction of
halos with a displacement $<d_{2D}^{\rm bar}$ can be considered as an
upper limit.

The results for the integrated distributions for $>d_{2D}^{\rm bar}$
are shown in Figure \ref{fig:baryonic_displacements}. The dashed lines
represented the results for groups and the continuous lines correspond
to clusters. As a test of our model we compare the cluster results
against the analytic fit provided by \cite{ForeroRomero2010}. This fit
reproduces the statistics for the DM-baryon separation found for
clusters more massive than $>10^{14}$\hMsun\ in a simulation
which included a description for DM and gas with $8$ times the volume of the
Bolshoi Simulation. The fit is valid for separations larger than
$70$\hkpc, beyond which we find that it provides a remarkably good
description within a factor of $\sim 2$ of our results. This gives us
confidence in our approach to estimate the expected fraction of
groups with a DM-baryon displacement.

From Figure \ref{fig:baryonic_displacements} we see that only a
fraction of $0.1\%$ to $1\%$ of the groups are expected to have a
DM-baryon displacement equal or larger than $87\pm 14$\hkpc as
observed in \bullg. This fraction rises to $4\%$ to $10\%$ in the case
of clusters, consistent with the results reported by
\cite{ForeroRomero2010}.

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{figure_4.eps}
\end{center}
\caption{Integrated probability distribution for the estimated
  baryonic displacements in the group and cluster samples. Continous (dashed)
  lines correspond to clusters (groups). The continuous black line
  marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and baryons. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.} 
\label{fig:baryonic_displacements}
\end{figure}

\subsection{Possible Observational Statistical Tests}

Recently \citep{Foex2013} presented an analysis of $80$ galaxy groups
in the SL2S sample. From the light distribution, only $34$ objects
($\sim 42\%$) have regular isophotes, $33$ had ellongated
isophotes (hints of merging system), and $13$ ($\sim 16\%$) had a
clear bi-modal light distribution.  The system \bullg\ is one of these
$13$ systems.

@@

We now make a comparison of these fractions against the results of our
simulations. The results are summarized in Table
\ref{table:numbers}. The first row indicate the total number of halos
in each sample, at each redshift. The second row shows the number of
objects with DM-DM displacement $46\hkpc< d_{\rm 2D}< 350$\hkpc. The
last row indicates the number of objects in the previous sub-sample
with a  $d_{\rm 2D}^{\rm bar}>87$\hkpc.

Considering that the statistics for the cluster sample are dominated
by objects in the mass range of the \bullg we make a comparison
against this sample. Table \ref{table:numbers} shows that $\sim 40\%$ of the 
clusters are expected to have large DM-DM displacements, this is a
factor of $\sim 2$ larger than the observational estimate by
\citep{Foex2013}. However, roughly $\sim 7\%$ of these systems present a
displacement equal or larger than the observed in the \bullg, a
fraction that is compatible with the $1/13\sim 0.07$ fraction in the
observational sample.

This rough comparison  that our estimates for the relative number of
bullet systems with respect to multi-modal systems is compatible with
observations. A proper comparison must take into account all the observational
uncertainties, biases, mixes between our two populations (groups \&
clusters at different redshifts) to derive a stronger prediction from
the simulation, something that is beyond the scope of this Letter.

Neverthless, comparing the group and cluster samples we expect to have
a number density of bullets similar to \bullg that is three times
larger for groups than for clusters.

>>>>

We estimate the expected distribution of displacements between the two
dominant dark matter density peaks and between baryons and dark matter
in halos simulated within a full cosmological context in $\Lambda$CDM.
We use as a benchmark the observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated
baryonic component. We extend to lower masses previous computational results by
studying two halo samples: groups and clusters. We find that $(50 \pm
10)\%$ of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$186 \pm 30$\hkpc as observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $(70 \pm 10)\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1$ to
$1.0\%$ of the groups should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for clusters this fraction rises to $(7\pm 3)\%$, consistent
with previous studies of dark matter to baryon separations. In terms
of absolute number densities we find that the expected value for
groups similar to \bullg\ is three times larger than the expected value
for clusters. This extension of the theoretical predictions and
observational results towards low mass bi-modal dark matter
configurations opens up the possibility for a new statistical test of
$\Lambda$CDM.   
\end{abstract}

@@

In this letter we work with two quantities that could be inferred
from observations of Bullet-like systems. The projected distance
between two dominant DM clumps, $d_{\rm 2D}$ and the projected distance
between the DM-baryonic clumps, $d_{\rm 2D}^{\rm bar}$.

From the simulation point of view, the first quantity can be
translated into the 2D projected values of $||\vec{r}||$ and its value
relative to the virial radius $D_{\rm   off}= ||\vec{r}||_{\rm
  2D}/R_{\rm vir}$. The second quantity, the projected DM-baryon
distance, is not directly available from a DM-only simulation but can
be estimated from the data.

@@

\subsection{DM-DM Displacements}
\label{fig:displacement}

@@

Figure \ref{fig:displacement} presents the integrated
probability distribution for the DM-DM displacements, $d_{\rm 2D}$ The
left panel shows the displacement in physical units and the right
panel as a fraction of the virial radius of the host halo.  The panel
with the projected 2D physical displacements also shows a vertical
stripe with the estimated displacement for the Bullet-group reported
by \cite{Gastaldello}.

In the group sample we see that a fraction of
$40\%$ to $60\%$  should present a displacement equal than the
estimate for \bullg; in the cluster sample this fraction increases to
$70\%$-$80\%$. The uncertainties in this estimates are derived from
the uncertainties in the displacement measurement for \bullg. This
fraction is naturally higher in more massive systems because they are
larger in size. Normalizing the displacements by the virial radius,
right panel Figure \ref{fig:displacement}, we see that the
distributions are similar for the two samples at all redshifts.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{figure_2.eps}
\end{center}
\caption{2D histograms in the plane $\mu$-$D_{\rm off}$. The left
  panel corresponds to clusters and the right panel to groups. The data
  used to construct this histogram includes the halos at all redshifts.}
\label{fig:geometry}
\end{figure*}

Figure \ref{fig:geometry} presents the geometry of the intereactions using the 
variables $\mu$ and $D_{\rm off}$. The first evident feature is that
most of the configurations have head-on encouters, $|\mu|>0.9$
($\theta\leq 30^{\circ}$), while only a minority with $|\mu|<0.9$ can
be described as having  grazing trajectories.

For pairs on radial trajectories there are three regions of interest
in this plane that correspond to the three merging stages described
at the end of Section \ref{sec:geometry} assuming that the sub-halo
merges (or falls below the BDM detection threshold) right at its
second pass through the center of the host halo \citep{Poole2006}.

The first region has $\mu\approx-1$ and $D_{\rm off}>0.6$, which
locates the systems where a head-on collision has just started. The
second region has $\mu\approx 1$ and $D_{\rm off}<0.6$; at this stage
the collision continues after the first crossing of the host's center,
the low number of halos with radial infalling velocities and
displacements $D_{\rm off}>0.6$ suggest that this the maximum range of
radii for the apogee.  The third region corresponds to $\mu\approx-1$
and $D_{\rm off}<0.6$ that is the the secondary infall after apogee.

In the next subsection we use the information in this sequence to
estimate the expected DM-baryon displacement.

\subsection{DM-baryon Displacement}
\label{sec:baryonic_displacements}

The results we have derived so far apply to multi-modal systems and
the expected separation between the two dominant dark matter
clumps. However, a non-zero DM-DM displacement does not imply 
a non-zero DM-baryon displacement. Given that we use a DM only
simulation, we have to find a way to estimate a DM-baryon displacement
from the available information.

This estimate is based on the different kinds of trajectories and
collision stages described in previous sections. To start with, we
consider that systems with $|\mu|<0.9$, describing grazing
trajectories, have a baryonic displacement equal to zero.  In systems
with head-on collissions, $|\mu|>0.9$, we distinguish two different
cases. The first one, systems with  $\mu<-0.9$ and large displacements $D_{\rm
  off}>0.6$ most probably describe the beginning of the interaction
and should have baryonic displacements equal to zero.
equal to zero. In all the other cases, the DM-baryon displacement
should be equal to zero. We approximate this displacement by  $d_{\rm
  2D}^{\rm bar} = X_{\rm off} R_{\rm   vir}$, where $X_{\rm   off}$ is
the offset computed between the minium of potential and the center of
mass for each host halo computed from all the matter inside the virial
radius.

This simplified model does not take into account that there is a
fraction of halos with $\mu<-0.9$ and $D_{\rm off}<0.6$ for which the
collision has not started and should have $d_{2D}^{\rm bar}=0$. A detailed
modeling of this fraction requires the study of the complete merger
tree of the halo and sub-halo, a study beyond the scope of this
Letter. Instead we caution the reader that the derived fraction of
halos with a displacement $<d_{2D}^{\rm bar}$ must be considered as an
upper limit.

The results for the integrated distributions for $>d_{\rm 2D}^{\rm bar}$
are shown in Figure \ref{fig:baryonic_displacements}. The dashed lines
represented the results for groups and the continuous lines correspond
to clusters. As a test of our model we compare the cluster results
against the analytic fit provided by \cite{ForeroRomero2010}. This fit
reproduces the statistics for the DM-baryon separation found for
clusters more massive than $>10^{14}$\hMsun\ in a simulation
which included a description for DM and gas with $8$ times more volume
and $8$ times less mass resolution than the Bolshoi Simulation. The
fit is valid for separations larger than $70$\hkpc, beyond which we
find that it provides a remarkably good description within a factor of
$\sim 2$ of our results. This gives us confidence in our approach to
estimate the expected fraction of groups with a DM-baryon
displacement.

Finally, from Figure \ref{fig:baryonic_displacements} we see that only a
fraction of $0.1\%$ to $1\%$ of the groups are expected to have a
DM-baryon displacement equal or larger than $87\pm 14$\hkpc as
observed in \bullg. This fraction rises to $4\%$ to $10\%$ in the case
of clusters, consistent with the results reported by
\cite{ForeroRomero2010}.

\subsection{Towards a Statistical Comparison Against Observations}

Recently \cite{Foex2013} presented an analysis of $80$ galaxy groups
in the SL2S sample. From the light distribution, only $34$ objects
($\sim 42\%$) have regular isophotes, $33$ had ellongated
isophotes (hints of merging system), and $13$ ($\sim 16\%$) had a
clear bi-modal light distribution; \bullg\ is one of these
$13$ systems.

@@

We now make a comparison of these fractions against the results of our
simulations. The results are summarized in Table
\ref{table:numbers}. The first row indicates the total number of halos
in each sample at each redshift. The second row shows the number of
objects with DM-DM displacements $46\hkpc< d_{\rm 2D}< 350$\hkpc. The
last row indicates the number of objects in the previous sub-sample
with DM-baryon displacements $d_{\rm 2D}^{\rm bar}>87$\hkpc.

Considering that the statistics for the cluster sample are dominated
by objects in the mass range of the \bullg we make a comparison
against this sample. Table \ref{table:numbers} shows that $\sim 40\%$ of the 
clusters are expected to have large DM-DM displacements, this is a
factor of $\sim 2$ larger than the observational estimate by
\cite{Foex2013}. However, roughly $\sim 7\%$ of these systems present a
displacement equal or larger than the observed in the \bullg, a
fraction that is compatible with the $1/13\sim 0.07$ fraction in the
SL2S sample from which \bullg\ was drawn.

This rough comparison  shows that our estimates for the relative number of
bullet systems (large DM-baryon displacements) with respect to
multi-modal systems (large DM-DM displacements) is compatible with
observations. A proper comparison must take into account all the observational
uncertainties, biases, mixes between our two populations (groups \&
clusters at different redshifts) to derive a stronger bound from
the simulation, something that is beyond the scope of this Letter.

Neverthless, comparing the number for group and cluster samples in the
last row of Table \ref{table:numbers} we can already predict that a
fixed redshift the number density of bullets similar to \bullg that
three times larger for groups than for clusters.

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{figure_3.eps}
\end{center}
\caption{Integrated probability distribution for the estimated
  baryonic displacements in the group and cluster samples. Continous (dashed)
  lines correspond to clusters (groups). The continuous black line
  marked as $P_{2D}$ shows the statistics reported by
  \citet{ForeroRomero2010} for a cosmological simulation including DM
  and baryons. The vertical lines correspond to the mean value and
  uncertainty of the displacement measured for \bullg.} 
\label{fig:baryonic_displacements}
\end{figure*}
